---
title: "code_portfolio"
author: "Nicasio"
date: "April 1, 2019"
output: 
  html_document:
    toc: True
---

# Week 2

[R Tutorial](http://www.r-tutor.com/r-introduction)

Variable Assignment
```{r}
x = 1
x
```

Functions
```{r}
# c() is the combine function
c(1, 2, 3)
```


Getting Help

```{r}
# you can use ?function or help(function)
?c
```


## Basic Data types

### Numeric

```{r}
x = 10.5  # assign a decimal values
x         # print the value of x
class(x)  # print the class name of x
```

`Numeric` is not the same as an `integer`
```{r}
k = 1
k
class(k)
is.integer(k) # is k an integer? is.integer function
```

### Integer

How to create an `integer`
```{r}
y = as.integer(3)
y
class(y)
is.integer(y)
```
Another way to create `integer`
```{r}
y = 3L #note adding the string "L"
is.integer(y)
```

You can convert `numeric` values, decimal strings, and TRUE/FALSE `logical` values

```{r}
as.integer(3.64)  #coerce numeric value
as.integer("5.67") #coerce decimal string
as.integer(TRUE) #coerce logical values
```
* Note that values are rounded DOWN to the whats left of the decimal

You can't convert non-decimal strings
```{r}
as.integer("joe")
```

### Complex

A **complex** value in R is defined via the pure imaginary value of *i*.

```{r}
z = 1 +2i #create a complex number
z
class(z)
```

### Logical

```{r}
x = 1; y = 2  #sample values
z = x > y     # is x greater than y? assign to variable z
z
class(z)
```

### Character

```{r}
x = as.character(3.14)
x
class(x)
```

Two character values can be concantenated with the `paste` function
```{r}
fname = "Joe"; lname = "Smith"
paste(fname, lname)
```

`sprintf` function uses C language syntax

```{r}
sprintf("%s has %d dollars", "Sam", 100)
```

Extract substring with `substr` function

```{r}
substr("Mary has a little lamb.", start = 3, stop = 12)
#start and stop uses index
```

replace a word in the character value, use `sub` function
```{r}
sub("little", "big", "Mary has a little lamb.")
```

## Vector

A **vector** is a sequence of data elements of the same basic type

```{r}
X <- c(2, 3, 4)   #vector of numeric values
Y <- c(TRUE, FALSE, TRUE)   #vector of logical values
Z <- c("a", "b", "c")     #vector of characters
```

You can use `length` function to get the count of values in the vector
```{r}
length(Z)
```

### Combine Vectors
```{r}
c(X,Y,Z)
```

### Vector Arithmetics
```{r}
a = c(1, 3, 5, 7)
b = c(1, 2, 4, 8)

5 * a #multiply each number in vector a by 5
a + b #adding vectors.
a - b #subtracting vectors
a * b #multiply vectors
a / b #divide vectors
```

Recycling rule. If vectors are not of equal length then the the shorter vector will be "recycled" or used again for the rest of the larger vector.
```{r}
u = c(10, 20, 30)
v = c(1 , 2, 3, 4, 5, 6, 7, 8, 9)
u + v
```

### Vector index

We've seen earlier with `substr` function how vectors have an index based on the position of each value.

```{r}
s = c("aa", "bb", "cc", "dd", "ee") 
s[3]
```

You can delete using a negative symbol index
```{r}
s[-3]
```

You can use `c` function to grab multiple values from a `vector`
```{r}
s[c(2,3)] #gets 2nd and 3rd value from s vector
s[c(2,3,3)] #gets 2nd, 3rd, and duplicate of 3rd value
s[c(2,1,3)] #you can index out of order
```

Range index is used to slice a chunk of the vector based on a given range.
```{r}
s[2:4] #gets 2nd, 3rd, and 4th value of vector s
```

### Logical index

You can use `logical` values to index
```{r}
L = c(FALSE, TRUE, FALSE, TRUE, FALSE)
s[L]
```

### Named Vectors

You can assign names to a vector using `names` function
```{r}
v = c("Mary", "Sue")
names(v) = c("First", "Last")
v["First"]
```

## Matrix

A **matrix** is a collection of data elemets arranged in a two-dimensional rectangula layout. You can use the `matrix` function to make a matrix

```{r}
A = matrix(
  c(2, 4, 3, 1, 5, 7), #the data elements
  nrow = 2,            #number of rows
  ncol = 3,            #number of columns
  byrow = TRUE         #fill matrix by rows
)
A
```

You can index a matrix by *m*^th^ row and *n*^th^ column using A[*m*,*n*] expression
```{r}
A[2,3]
```

If you leave *n* empty it will return the entire row
```{r}
A[2,]
```
If you leave *m* empty it will return the entire column
```{r}
A[,3]
```

You can use `c` function to return more than one row or column
```{r}
A[,c(1,3)]
```

You can assign names to rows and columns
```{r}
dimnames(A) = list(
  c("row1", "row2"),  #row names
  c("col1", "col2", "col3")  #column names
)
A
```

You can now index by row and column name
```{r}
A["row2", "col3"]
```

### Matrix Construction

The different ways to construct a matrix
```{r}
#classic way
B = matrix(
  c(2, 4, 3, 1, 5, 7),
  nrow=3,
  ncol=2)

B
```

You can transpose a matrix using `t` function
```{r}
#transpose
t(B)
```

Combining matrices by column using `cbind` function
```{r}
C = matrix(
  c(7,4,2),
  nrow = 3,
  ncol = 1
)

cbind(B, C) #combine matrix B and C
```

Combining matrices by row using `rbind` function
```{r}
D = matrix(
  c(6,2),
  nrow = 1,
  ncol = 2
)

rbind(B,D)
```

You can deconstuct (or flatten) a matrix using `c` function
```{r}
c(B)
```

## List

A **list** is a like a vector but it contains different `class` objects

```{r}
n = c(2,3,5)
s = c("aa", "bb", "cc", "dd", "ee")
b = c(TRUE, FALSE, TRUE, FALSE, FALSE)
x = list(n, s, b, 3)   # x contains copies of n, s, b
x
class(x)
```

List Slicing uses a *single square bracket* "[]"
```{r}
x[2] #single member of list

x[c(2,4)] #multiple members of a list
```

Member reference
using **double square brackets** "[[]]" will return a copy of the vector in the list.

```{r}
x[[2]] #returns a copy of s vector
class(x[[2]]) #this is a character vector
class(x[2]) #this is a slice of a list
```

You can modify a vector within a list
```{r}
x[[2]][1] = "ta"
x[[2]]
s # s is unaffected
```

### Named list members

Assigning names to list members
```{r}
v = list(bob = c(2,3,4), john = c("aa", "bb"))
v
```
List slicing with names
```{r}
v["bob"]
```
List slicing multiple names
```{r}
v[c("bob","john")]
```

Member reference
```{r}
v[["bob"]] #note double brackets

v$bob #note $ symbol
```

Search Oath. You can *attach* a list to the R search path and access its members without explicitly mentioning the list. It should be *detached* for cleanup
```{r}
attach(v)
bob
detach(v)
```

## Data Frame

A **data frame** is used for storing data tables. It is a list of vectors of equal length. For example, the following variable `df` is a data frame containing three vectors `n`,`s`,`b`.

```{r}
n = c(2,3,5)
s = c("aa", "bb", "cc")
b = c(TRUE, FALSE, TRUE)
df = data.frame(n, s, b)
```

Built-in Data Frame

```{r}
mtcars
```

Indexing a Data Frame
```{r}
mtcars [1,2]  #cell value first row, second column
mtcars["Mazda RX4", "cyl"] #index by name
```

Preview the Data Frame
```{r}
nrow(mtcars)  #returns number of rows of data frame
ncol(mtcars)  #returns number of columns
head(mtcars)  #preview the top of the Data Frame
```

```{r}
help(mtcars)
```

### Data Frame Column Vector

```{r}
mtcars[[9]] #by index number
mtcars[["am"]]  #by name
mtcars$am  #using $
mtcars[,"am"] #list index
```

### Data Frame Column Slice

```{r}
mtcars[1] #numeric indexing
mtcars["mpg"] #name indexing
mtcars[c("mpg", "hp")] #multiple column slice
```

### Data Frame Row Slice

```{r}
mtcars[24,] #numeric indexing
mtcars[c(3, 24),] #multiple numeric indexing
mtcars["Camaro Z28",] #name indexing
mtcars[c("Datsun 710", "Camaro Z28"),] #multiple name indexing
```

```{r}
L = mtcars$am == 0 #Logical indexing
mtcars[L,] #list of vehicles with automatic transmissions
mtcars[L,]$mpg #gas mileage for automatic tranmission cars
```

### Data Import

**Excel File**

```
library(gdata)  #load gdata package
mydata = read.xls("mydata.xls")  #read from first sheet
```

```
library(XLConnect)  # load XLConnect package
wk = loadWorkbook("mydata.xls") 
df = readWorksheet(wk, sheet="Sheet1")
```

**Minitab File**

```
library(foreign)                 # load the foreign package 
mydata = read.mtp("mydata.mtp")  # read from .mtp file
```

**SPSS File**

```
library(foreign)                 # load the foreign package 
help(read.spss)                  # documentation 
mydata = read.spss("myfile", to.data.frame=TRUE)
```

**Table File**

```
mydata = read.table("mydata.txt")  # read text file 
mydata                             # print data frame 
```

**CSV File**

```
mydata = read.csv("mydata.csv")  # read csv file 
mydata 
```


**Working Directory**
Finally, the code samples above assume the data files are located in the R working directory, which can be found with the function getwd.

`getwd()               # get current working directory`

You can select a different working directory with the function setwd(), and thus avoid entering the full path of the data files.

`setwd("<new path>")   # set working directory`

Note that the forward slash should be used as the path separator even on Windows platform.

`setwd("C:/MyDoc")`